Network Structure, Efficiency, and Performance in WikiProjects

### 2017-08-08 paper.tex
Add figures, table formatting, related work.

% Document-specific packages
\usepackage{epigraph}
\usepackage[flushleft]{threeparttable}
\usepackage{graphicx}

@@

Lab-based experiments on networked collaboration
suggest a complex interaction between network topology,
and other factors.
While groups of networked human subjects reliably perform very well on
difficult graph-coloring tasks, the best performing network architectures
(e.g., fully-connected vs. small-world) vary
from task to task \cite{kearns_experiments_2012}.
The same studies found that while human subjects tend to perform very well on
a number of networks, they perform worst on collaboratively self-organized
networks, highlighting the importance of intentionality
in constructing communication channels for collective decision-making.
Similarly, some network topologies are able to reach faster decisions in the
presence of more information, while others show the opposite effect
\cite{kearns_experimental_2006}.
Based on lab experiments, Fowler and Christakis \cite{fowler_cooperative_2010}
suggest that individual decisions towards altruism are conditional on their
neighbor's behavior and ``contagious'' up to three degrees away.
Later experiments by Suri and Watts \cite{suri_cooperation_2011} confirmed the
existence of conditional altruism,
but conlcuded that altruistic
behavior only influences first-degree neighbors.

Collaboration as decision-making
Collaboration as consensus

Collaboration as community
Collaboration and social capital

Collaboration on Wikipedia

This project

@@

OLD STUFF BELOW HERE

Collaboration has also been studied extensively through the lens of collective
decision-making and voting theory.
The Condorcet jury theorem and its extensions \cite{list_epistemic_2001}
describe how individuals can use majority voting to improve their
decision-making ability.
The process of decision-making is sometimes divided into
``talking'' and ``voting'' stages.
While heirarchical representative democracy sometimes improves the outcome
of ``talk,'' it weakens the result of the Condoret jury theorem for voting
\cite{list_epistemic_2001}.
Robert Michels proposed the ``iron law of oligarchy,''
\cite{michels_political_1999} which states that
the earlier members of a group will, over time, gain disproportionate
decision-making power and act increasingly out of self-interest rather than
the good of the group.
Shaw and Hill found that behavior in online wiki communities is consistent
with the iron law \cite{shaw_laboratories_2014}.

A number of studies, summarized in \cite{gentry_consensus_1982},
have examined consensus-based decision-making procedures, used extensively by
the Quakers and later by activist communities.
The hallmark of consensus is the equal and active participation by all members
of a group.
Key findings include the tradeoff between speed and quality of decisions,
the importance of expressing conflict early in the process,
and the willingness to recognize when one's own views are counter to the
existing consensus.
The details of why this tradeoff occurs remain an open question.
De Tar created InterTwinkles \cite{detar_intertwinkles:_2013},
an online suite of tools to facilitate consensus decision-making
in small cooperatives.
With a better understanding of how large-scale network structure influences
decision-making, such tools could be extended to large collaborations.

Research on digital communities has also examined some non-network factors
influcencing the quality of decisions and collaborative work.
Using a simulation model, Hong and Page \cite{hong_groups_2004} found that
diverse groups can outperform groups composed of the best individual
problem-solvers.
A similar result was observed on Wikipedia by
Robert and Romero, who found that
larger group sizes yield higher article ratings
(as determined by editor peer-review)
when the groups are diverse and experienced
\cite{robert_when_2015}.
Also on Wikipedia,
Kittur and Kraut found that different types of coordination have a complex
effect on the quality of Wikipedia articles \cite{kittur_harnessing_2008}.
Both explicit and implicit coordination result in higher quality articles,
with explicit coordination being especially central in the early life of an
article.

In sociology and anthropology, it has often been observed that an individual's
position within a social network can result in privilege or power relative to
others.
For example, powerful individuals are often ``brokers''
who act as exclusive intermediaries between disconnected portions of the
social network \mbox{\cite{silverman_patronage_1965}}.
Similarly, successful innovation in organizations often occurs in ``structural
holes'' between groups \mbox{\cite{granovetter_strength_1973}}.
Thus, the structure of a social network might encode meaningful information
about the distribution of power and social capital within a group.
This \mbox{\em strucural (in)equality} might have important consequences for the
ability of a group to collaborate.
There is some evidence that groups with high structural inequality,
as evidenced by highly-skewed degree distributions,
perform worse on colaborative tasks \mbox{\cite{kearns_experiments_2012}}.

Across the broad range of work on collaboration and related topics,
a few key themes emerge.
The efficiency and performance of collaborations are important considerations
and vary between different networks and problem types.
While social learning predicts no relationship between the two,
contagion-style innovation models predict a tradeoff.
Such a tradeoff has been observed in networked problem-solving and in
different forms of group decision-making.

In this project, I will investigate several gaps in the existing literature.
In most cases, degree distribution seems to correlate to performance.
But aside from the naive Bayes case in social learning, it is unkown whether
the correlation is explained best by degree or by another structural property
correlated with degree.
The characteristic path length and min-cut studied here are candidates for such
degree-correlated factors.
While existing work has focused on families of artificial networks,
this project examines a large number of naturally-occuring networks.
Unlike artificial networks, the structural properties of these naturally-occuring
networks can vary independently, making it easier to isolate individual
network properties that correlate with outcome variables.
This project will also test whether the detrimental effects of skewed
distributions of network properties that have been observed in lab experiments
\mbox{\cite{kearns_experiments_2012}}
can be observed in large-scale, naturally-occuring data.

@@

>>>>

% Document-specific packages
\usepackage{epigraph}
\usepackage[normal]{threeparttable}
\usepackage{graphicx}

@@

Lab-based experiments on networked collaboration
suggest a complex interaction between network topology,
and other factors.
While groups of networked human subjects reliably perform very well on
difficult graph-coloring tasks, the best performing network architectures
(e.g., fully-connected vs. small-world) vary
from task to task \cite{kearns_experiments_2012}.
The same studies found that while human subjects tend to perform very well on
a number of networks, they perform worst on collaboratively self-organized
networks, highlighting the importance of intentionality
in constructing communication channels for collective decision-making.
Similarly, some network topologies are able to reach faster decisions in the
presence of more information, while others show the opposite effect
\cite{kearns_experimental_2006}.
Based on lab experiments, Fowler and Christakis \cite{fowler_cooperative_2010}
suggest that individual decisions towards altruism are conditional on their
neighbor's behavior and ``contagious'' up to three degrees away.
Later experiments by Suri and Watts \cite{suri_cooperation_2011} confirmed the
existence of conditional altruism,
but conlcuded that altruistic
behavior only influences first-degree neighbors.
While theories of networked social learning have been tested in small-scale lab experiments,
there is a need to study real-world, large-scale collaborations,
as we do in the present paper.

In addition to learning, decision-making is a crucial component of collaboration.
The Condorcet jury theorem and its extensions \cite{list_epistemic_2001}
describe how individuals can use majority voting to improve their
decision-making ability.
The process of decision-making is sometimes divided into
``talking'' and ``voting'' stages.
While heirarchical representative democracy sometimes improves the outcome
of ``talk,'' it weakens the result of the Condoret jury theorem for voting
\cite{list_epistemic_2001}.
Robert Michels proposed the ``iron law of oligarchy,''
\cite{michels_political_1999} which states that
the earlier members of a group will, over time, gain disproportionate
decision-making power and act increasingly out of self-interest rather than
the good of the group.
Shaw and Hill found that behavior in online wiki communities is consistent
with the iron law \cite{shaw_laboratories_2014}.
A number of studies, summarized in \cite{gentry_consensus_1982},
have examined consensus-based decision-making procedures, used extensively by
the Quakers and later by activist communities.
The hallmark of consensus is the equal and active participation by all members
of a group.
Key findings include the tradeoff between speed and quality of decisions,
the importance of expressing conflict early in the process,
and the willingness to recognize when one's own views are counter to the
existing consensus.
De Tar created InterTwinkles \cite{detar_intertwinkles:_2013},
an online suite of tools to facilitate consensus decision-making
in small cooperatives.
Consensus decision making is notoriously difficult to apply in large groups.
Wikipedia is one of the largest collaborations to use consensus-based
decision-making,
making it an excellent place to study the role of network structure in
large-scale collaboration.

Research on digital communities has also examined some non-network factors
influcencing the quality of decisions and collaborative work.
Using a simulation model, Hong and Page \cite{hong_groups_2004} found that
diverse groups can outperform groups composed of the best individual
problem-solvers.
In sociology, much research has been done examining the relationship between
network structure and social capital.
Powerful individuals are often ``brokers''
who act as exclusive intermediaries between disconnected portions of the
social network \mbox{\cite{silverman_patronage_1965}}.
Similarly, successful innovation in organizations often occurs in ``structural
holes'' between groups \mbox{\cite{granovetter_strength_1973}}.
There is some evidence that groups with high structural inequality,
as evidenced by highly-skewed degree distributions,
perform worse on colaborative tasks \mbox{\cite{kearns_experiments_2012}}.
In the present project, we examine the effect of structural inequality
on collaboration in Wikipedia.

Several studies have examined the role of community in the specific context
of Wikipedia.
Robert and Romero found that
larger group sizes yield higher article ratings
(as determined by editor peer-review)
when the groups are diverse and experienced
\cite{robert_when_2015}.
Also on Wikipedia,
Kittur and Kraut found that different types of coordination have a complex
effect on the quality of Wikipedia articles \cite{kittur_harnessing_2008}.
Both explicit and implicit coordination result in higher quality articles,
with explicit coordination being especially central in the early life of an
article.

Across the broad range of work on collaboration and related topics,
a few key themes emerge.
The efficiency and performance of collaborations are important considerations
and vary depending on both network structure and type of task.
While social learning predicts no relationship between the two,
contagion-style innovation models predict a tradeoff.
Such a tradeoff has been observed in simulations and lab experiments on
collaboration.
In this project, I will investigate several gaps in the existing literature.
In much of the existing literature,
degree distribution correlate to performance.
But aside from the naive Bayes case in social learning, it is unkown whether
the correlation is explained best by degree or by another structural property
correlated with degree.
The characteristic path length and min-cut studied here are candidates for such
degree-correlated factors.
While existing work has focused on families of artificial networks,
this project examines a large number of naturally-occuring networks.
Unlike artificial networks, the structural properties of these naturally-occuring
networks can vary independently, making it easier to isolate individual
network properties that correlate with outcome variables.
This project will also examine Wikipedia for evidence that structural
inequality (in the form of skewed
degree distributions) has a deterimental effect on performance,
as has been observed in lab experiments
\mbox{\cite{kearns_experiments_2012}}.

@@

@@

\begin{figure*}
\includegraphics[width=6.67in,height=1.33in]{fig-deg-eff.pdf}
\includegraphics[width=6.67in,height=1.33in]{fig-deg-perf.pdf}
\caption{
TODO
\label{fig:deg-eff-perf}
}
\end{figure*}

\begin{table}
\small
\caption{
Efficiency regression coefficients.
\label{tab:sim-eff}
}
\centering
\bigskip
\begin{tabular}{llc}
\hline
Strategy & Std. Coeff. \\
\hline
Best-neighbor       & -0.064 \\
Conformity          & -0.080 \\
Local best-neighbor & -0.086 \\
Local conformity    & \+0.049 \\
Consensus           & -0.0325$^{***}$ \\
\hline
\end{tabular}
\begin{tablenotes}
\item \centering * $p < 0.05$. ** $p < 0.01$. *** $p < 0.001$.
\end{tablenotes}
\end{table}

\begin{table}
\small
\caption{
Performance regression coefficients.
\label{tab:sim-perf}
}
\centering
\bigskip
\begin{tabular}{llc}
\hline
Strategy & Std. Coeff. \\
\hline
Best-neighbor       & \+0.084 \\
Conformity          & \+0.130 \\
Local best-neighbor & -0.080 \\
Local conformity    & -0.020 \\
Consensus           & \+2.38$\times{10^{-4\;**}}$ \\
\hline
\end{tabular}
\begin{tablenotes}
\item \centering * $p < 0.05$. ** $p < 0.01$. *** $p < 0.001$.
\end{tablenotes}
\end{table}
